---
title: "R Glossary for RDSVS MSc's"
author: "RDSVS R Team"
date: "2018-05-25"
output: 
  html_document: 
    highlight: haddock
    theme: spacelab
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

# Introduction
<span style="color:red">Some introduction and use text</span>

Where possible we'll include examples that you can run with R's built in data files.

We'll show that examples can be run by <span style="color:red"> how can we show that they are runnable?</span>



# Basics

### <-
___assign___

Read this as 'gets' so the following code to put create an object with the name ```num``` reads as `num gets 99`...

```{r, eval=FALSE}
num <- 99
```
-----

### print()
___print___

This function 'prints' out its arguments to the console or document. Code typed into the console will print automatically so the following give the same result...

```{r}
print(999)
```

```{r}
999
```
-----

### c()
___combine___

This function combines it's arguments to make a single set or in R language a 'vector'...

```{r}
numbers <- c(1, 4, 6, 2)
numbers
```
-----

### %>% 
___the pipe___

This operator takes the output of the code before it and 'pipes' it into the first argument of the next function. So the following two chunks of code do the same thing...

```{r}
x <- c(1, 5, 6, 7)
x <- mean(x)
print(x)
```

(You need to load the 'tidyverse' to use the pipe)
```{r, include=FALSE}
library(tidyverse)
```

```{r, eval = FALSE}
library(tidyverse)
```


```{r}
c(1, 5, 6, 7) %>% 
  mean() %>% 
  print()
```




# Data import

The easiest way, at first, to import data is to use RStudio's helpers in the environment tab. Here you can click on 'Import Dataset' and then find the file you want to import, set some options and RStudio will generate the R code that you need to copy into your R script. Later you may want to just write the import code directly - it's not too hard. Here are the main functions you'll use...

### read_excel
___Import an Excel spreadsheet___

This function needs the '''readxl''' library to be loaded. You give it the filename (or path), the sheet name (or number) and optionally other arguments such as the number of lines to skip, inforamtion about column names and types.

```{r, eval = FALSE}
mydata <- read_excel("my_excel_spreadsheet.xlsx",
                     sheet = "main-results")
```


# Data tidying

### rename()
___rename colums in a data.frame___

Imported data may have column names that are overly long, messy or contain characters that upset R. You can refer to 'upsetting' columns by putting backticks (`) around them but it's neater to rename columns to make them sensible. So if a data.frame, ```mydata``` had columns including ```Glucose level(mg/L)``` and ```silly_long_name_thats_a_pain_to_type``` we can rename them like this...

```{r, eval=FALSE}
mydata <- mydata %>% 
  rename(glucose = `Glucose level(mg/L`,
         better = silly_long_name_thats_a_pain_to_type)
```



# Data manipulation

# Plotting

# summary statistics

# Statistical tests

# Statistical models
