---
title: "R Glossary for RDSVS MSc's"
author: "RDSVS R Team"
date: "2018-05-25"
output: 
  html_document: 
    highlight: haddock
    theme: spacelab
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

# Introduction
<span style="color:red">Some introduction and use text</span>

Where possible we'll include examples that you can run with R's built in data files.

We'll show that examples can be run by <span style="color:red"> how can we show that they are runnable?</span>

We'll include links to further resources where approporiate. Including [R for Data Science](http://r4ds.had.co.nz) which is available online and <span style="color:red">Hadley's ggplot book or maybe the ggplot2 website would be better</span>. 

# Basics

### <-
___assign___

Read this as 'gets' so the following code to put create an object with the name ```num``` reads as `num gets 99`...

```{r, eval=FALSE}
num <- 99
```
-----

### print()
___print___

This function 'prints' out its arguments to the console or document. Code typed into the console will print automatically so the following give the same result...

```{r}
print(999)
```

```{r}
999
```
-----

### c()
___combine___

This function combines it's arguments to make a single set or in R language a 'vector'...

```{r}
numbers <- c(1, 4, 6, 2)
numbers
```
-----

### %>% 
___the pipe___

This operator takes the output of the code before it and 'pipes' it into the first argument of the next function. So the following two chunks of code do the same thing...

```{r}
x <- c(1, 5, 6, 7)
x <- mean(x)
print(x)
```

(You need to load the 'tidyverse' to use the pipe)
```{r, include=FALSE}
library(tidyverse)
```

```{r, eval = FALSE}
library(tidyverse)
```


```{r}
c(1, 5, 6, 7) %>% 
  mean() %>% 
  print()
```


-----

# Data import

The easiest way, at first, to import data is to use RStudio's helpers in the environment tab. Here you can click on 'Import Dataset' and then find the file you want to import, set some options and RStudio will generate the R code that you need to copy into your R script. Later you may want to just write the import code directly - it's not too hard. Here are the main functions you'll use...

### read_excel
___Import an Excel spreadsheet___

This function needs the '''readxl''' library to be loaded. You give it the filename (or path), the sheet name (or number) and optionally other arguments such as the number of lines to skip, information about column names and types and codes for missing data. The following code reads in the "main-results" sheet from a spreadsheet called "my_excel_spreadsheet.xlsx" and treats "*" as issing data. Note that the file will need to be in the same folder/directory as your project. If not you'll need to include the "path" to the file in the filename section. 

```{r, eval = FALSE}
mydata <- read_excel("my_excel_spreadsheet.xlsx",
                     sheet = "main-results",
                     na = "*")
```



# Data tidying and manipulation

### rename()
___rename colums in a data.frame___

Imported data may have column names that are overly long, messy or contain characters that upset R. You can refer to 'upsetting' columns by putting backticks (`) around them but it's neater to rename columns to make them sensible. So if a data.frame, ```mydata``` had columns including ```Glucose level(mg/L)``` and ```silly_long_name_thats_a_pain_to_type``` we can rename them like this...

```{r, eval=FALSE}
mydata <- mydata %>% 
  rename(glucose = `Glucose level(mg/L`,
         better = silly_long_name_thats_a_pain_to_type)
```

-----

### select()
___select columns and change column order___

The ```select()``` verb lets us choose or drop columns in a dataframe and change their order.

To make a new dataset using just the model, year and city fuel consumption columns from the mpg dataset (from the tidyverse packages) we can do this (we'll assume you've loded the tidyverse already with ```library(tidyverse)```)...

```{r}
new_data <- mpg %>% 
  select(model, year, cty)
```

Check it worked...

<span style="color:red">Is this too verbose???</span>

```{r}
new_data
```



See [here in R4DS](http://r4ds.had.co.nz/transform.html#select-columns-with-select) for more details.

-----

# Summary functions

### length
___returns the length of an R object___

Note - it includes missing values...

```{r}
ages <- c(1, 4, 1, 6, 3)
length(ages)

heights = c(12, 15, 17, NA, NA, 19)
length(heights)
```

To count only non-missing data use ```sum(!is.na())```...

```{r}
heights = c(12, 15, 17, NA, NA, 19)
sum(!is.na(heights))
```

The ```!is.na()``` is counted as ```1``` for every value that isn't missing and ```sum``` adds them up.

-----

### sum()
___calculate the total of a numeric vector___

```sum()``` returns the arithmetic sum of its argument(s). If there is any missing data it will return ```NA``` unless told to remove missing data first with ```na.rm = TRUE```. To find the total length of R's built in Rivers dataset...

```{r}
sum(rivers)
```

And if there's missing data...

```{r}
nums <- c(1, 6, 3, NA, 6)
sum(nums)
sum(nums, na.rm = TRUE)
```

-----







# Data Summary

# Plotting

# Statistical tests

# Statistical models
